\vspace{-2cm} \LARGE Divergence-Free Smoothed Particle Hydrodynamics \\ \large Advanced Simulation and Visualization of Fluids in Computer Graphics

### 2016-12-14 report/paper.tex
filled in some background work

    % Bridson
    ~\cite{bridson}.

    % Write about Benders article

@@

% Should be mentioned early since it is used in mostly of the algorithms
\subsection{Kernel} \label{kernel}
    A kernel function is used to simulate how particle-particle interactions decrease with the distance between the current particle and its neighbours. In SPH simulations this is an approximation of the Gaussian kernel function. Different kernels have been tested in previous works i.e. the poly6 kernel, the spiky kernel and the cubic spline kernel. According to Bender et. al.~\cite{bender} the cubic spline kernel presented by Monaghan~\cite{monaghan} was used. The kernel is described by Equation~\ref{eq:kernel}, where $q(x)=\frac{\left \| x \right \|}{h}$, $x$ is the distance between the current particle and a neighbour particle and $h$ is the support radius for the kernel. Particles further away than the support radius will not affect the current particle.

@@

\subsection{Navier-Stokes}
    Navier-Stokes equations describes incompressible fluids and is therefore suitable for SPH simulations. The equations expressed in Lagrangian coordinates can be found in Equation~\ref{eq:nsDensity} and~\ref{eq:nsVelocity}.

    \begin{equation} \label{eq:nsDensity}
        \frac{D\rho}{Dt} = 0 \Leftrightarrow \bigtriangledown \cdot \textbf{v} = 0 
    \end{equation}

    \begin{equation} \label{eq:nsVelocity}
        \frac{D\textbf{v}}{Dt} = -\frac{1}{\rho} \bigtriangledown p + v \bigtriangledown ^2 \textbf{v} + \frac{\textbf{f}}{\rho}
    \end{equation}

    In the equations $\frac{D\rho}{Dt}$ and $\frac{D\textbf{v}}{Dt}$ denotes the derivate of the density $\rho$ and the velocity $\textbf{v}$ respectively. The pressure, kinematic viscosity and body forces are denoted by $p,v$ and $\textbf{f}$ respectively. The second part of Equation~\ref{eq:nsDensity} $\bigtriangledown \cdot \textbf{v} = 0$ induce that the velocity field is free from divergence. From the continuity equation $\frac{D\rho}{Dt} = -\rho \bigtriangledown \cdot \textbf{v}$ and the divergence-free condition it follows that the partial derivate of the density with respect to the time is zero, which implies the equivalence of the equations in Equation~\ref{eq:nsDensity}. From this is follows that the density, in theory, must stay constant over time which is the implication of an incompressible fluid. In practice the divergence-free condition is not sufficient to guarantee incompressibility in simulations. The numerical time integrations will contain numerical errors which will cause density deviations due to the volume compressions. To avoid this a second condition $\rho - \rho_0 = 0$ called \textit{constant density condition} must be fullfilled~\cite{bender}. Bridson~\cite{bridson} uses splitting on Equation~\ref{eq:nsVelocity} which induce $\frac{D\textbf{v}}{Dt} = -\frac{1}{\rho} \bigtriangledown p$ which will be used further on in the simulation. Section~\ref{divergenceSolver} and~\ref{densitySolver} give a deeper description of the solvers used in the project.

\subsection{Divergence solver} \label{divergenceSolver}

\iffalse
\begin{lstlisting}[language=C++, mathescape=true, escapeinside={(*}{*)}]
function correctDivergenceError
    for all particles $i$ do
        calculateDensityChange $\frac{Dp}{Dt}_i$
    for all particles $i$ do //Scale $\color{codegreen}a_i$
        $a_i = \frac{a_i}{\Delta t}$
    while (t < $t_{max}$) do //Start simulation loop
        adapt time step $\Delta{t}$ 
        for all particles $i$ do //Predict velocities $\color{codegreen}\mathbf{v}_i^*$
            $\mathbf{v}_i = \mathbf{v}_i + \Delta{t}\mathbf{F}_i^{adv}/m_i$
        correctDensityError($\alpha, \mathbf{v}_i$) //Fulfill $\color{codegreen}\rho^* - \rho_0 = 0$
        for all particles $i$ do //Update positions
            $x_i(t + \Delta{t}) = x_i(t) + \Delta{t}\mathbf{v}_i^*$
        for all particles $i$ do //Update neighbourhoods
            find neighbourhoods $N_i(t + \Delta{t})$
        for all particles $i$ do //Update $\color{codegreen}\rho_i$ and $\color{codegreen}\alpha_i$ 
            compute densities $\rho_i(0)$
            compute factors $\alpha_i(0)$
        correctDivergenceError($\alpha, \mathbf{v}^*$) //Fulfill $\color{codegreen}\frac{D\rho}{Dt} = 0$
\end{lstlisting}
\fi

The divergence-free solver ensures the fluid to be free of density change over time by enforcing $\frac{D\rho}{Dt} = 0$, which is equivalent to the Navier-Stokes divergence free condition given in equation %\ref{eq:navier}.

The solver will ensure that the divergence error is of an acceptable level defined by the user. The divergence error is suppressed by a combined symmetric pressure force $\vect{F_{j\leftarrow i}^p} $ over all particle $i$'s neighbours $j$, that is applied to the predicted velocity of particle $i$. The pressure force is determined by

\begin{equation} \label{eq:pressure}
\vect{F_i^p} = -\frac{m_i}{\rho_i}\nabla \vect{p_i},
\end{equation}

where the pressure gradient can be solved using the SPH formulation proposed by Ihmsen et al.~\cite{ihmsen}.

\begin{equation} \label{eq:dpdt}
\frac{D\rho_i}{Dt} = \sum_j m_j (\vect{v_i} - \vect{v_j})\nabla \vect{W_{ij}}
\end{equation}

\begin{equation} \label{eq:kappa}
\nabla \vect{p_i} = \kappa_i^v \nabla \rho_i = \kappa_i^v \sum_j m_j \nabla \vect{W_{ij}}
\end{equation}

Equation \ref{eq:dpdt} defines how the density change is computed and \ref{eq:kappa} computes the pressure gradient with the help of the stiffness parameter $\kappa_i^v$, where

\begin{equation}
\kappa_i^v = \frac{D\rho}{Dt} \cdot \frac{\rho_i}{\Delta t \cdot \alpha_i}
\end{equation}

and $\alpha_i$ is the scaling factor only dependant on the position explained in section \ref{section:alpha}. The stiffness parameters needs to be determined iteratively since neighboring particles depend on each other.
The velocity of particle $i$ is thus changed to:

\begin{equation} \label{eq:pressuretotal}
\vect{v_i^*}(t) = \vect{v_i^*}(t) - \Delta t \sum_j m_j (\frac{\kappa_i^v}{\rho_i} \cdot \frac{\kappa_j^v}{\rho_i}) \nabla \vect{W_{ij}}
\end{equation}

After the symmetric pressure force has been applied to the new velocity should the density change in equation \ref{eq:dpdt} be equal or close to zero. If the error is too large will the process be iterated until desired result have been achieved.

\subsection{Density solver} \label{densitySolver}
    This solver aims to minimize the density error caused by the numerical integration in the divergence solver. The density error is determined by comparing the actual density to the rest density, $\rho - \rho_0$. There are several density solvers presented in previous works, but the solver used in this project was presented by Bender et. al.~\cite{bender} which uses the fact that $\alpha_i$ has been calculated before the solver which reduces the computations in the solver. The solver will be active while the average of the predicted densities for every particle differs from the rest density with an error greater than a constant \textit{maxError}. For every particle a predicted density $\rho_i^*$ is calculated through an Euler integration step by Equation~\ref{eq:predDensity}.

    \begin{equation} \label{eq:predDensity}
        \resizebox{0.88\linewidth}{!}{$
        \begin{matrix}
        \rho_i^* = \rho_i + \Delta t \frac{D\rho_i}{Dt} \\ 
        \\
        \Leftrightarrow \\ 
        \\
        \rho_i^* = \rho_i + \Delta t + \sum_j m_j 
        (\textbf{v}_i^* - \textbf{v}_j^*) \bigtriangledown W_{ij}
        \end{matrix}
        $}
    \end{equation}

    Since the stiffness parameter $\kappa_i$ is derived from the pressure forces, updating the stiffness parameter will correct the density error. The stiffness parameter is determined by Equation~\ref{stiffnessDensity}.

    \begin{equation} \label{stiffnessDensity}
    \kappa_i = \frac{1}{\Delta t^2}(\rho_i^*-\rho_0)\alpha_i
    \end{equation}

    The velocity is updated according to the the divergence solver. %Det här behöver kanske utvecklas, väntar med det tills stycket om divergence solvern är klar, innehåller samma ekvationer delvis

% Write about non pressure forces, velocity osv..?
\subsection{Adapted time step}

% CFL condition, explain why its necessary and how its assured that each rendered frame has the same time step.
In order to assure a stable simulation it is necessary to adapt the time step depending on the motion. 
The quicker the particles move the smaller time steps are taken such that the numerical 
error of the explicit time integration is negligible. For this is a condition proposed by 
Courant$-$Friedrichs$-$Lewy used often refered as the CFL-condition show in equation \ref{eq:cfl}. 
The time step $\Delta t$ is equal to a fraction of the particle's diameter $d$ divided by 
the maximum velocity $\vect{V_{max}}$ in the scene.

@@

\subsection{Density and alpha factors} \label{section:alpha}
    The density in a region of the fluid with a particle $x_i$ in the center is calculated through Equation~\ref{eq:density} where $\rho_i$ is the density with the current particle $x_i$ in center, $\rho_j$ is the density with a neighbouring particle $x_j$ in center, $m_j$ is the mass for a neighbour particle and $W_{ij}$ is the kernel function described in Section~\ref{kernel}. This implies that neighbouring particles closer to the current particle $x_i$ will have a greater affect on the density.

    \begin{equation}\label{eq:density}
        \rho_i = \sum_j \frac{m_j}{\rho_j} \rho_j W_{ij} = \sum_j m_j W_{ij}
    \end{equation}

\subsection{Screen space fluid rendering}

>>>>

    % Write about divergence free 
    In the field of computer graphics a variety of methods for simulating incompressible fluids exist.
    This section will give a brief overview of related approaches.

    Bender et al.~\cite{bender} proposed a method for a stable implicit SPH solution.
    The method uses a combination of two pressure solvers which enforce a low volume compression below 0.01$\%$ and a divergence-free velocity field.
    It can be seen as enforcing incompressibility on both position and velocity level.
    The low compression is important for realistic physical behavior and a divergence-free state increases the stability of the simulation which reduces the number of solver iterations.

    Another incompressible SPH method was proposed by Hu et al.~\cite{hu} for multiphase flows where the time integration step to obtain velocities is divided into two half steps.
    This is done to enforce both zero density variaton condition and the velocity divergence-free condition at each full time step.
    During the first half step density fluctuations are eliminated by altering the intermediate particle position.
    During the second half step errors in velocity divergence are resolved by altering the intermediate particle velocity.

    Ihmsen et al.~\cite{ihmsen2014implicit} proposed an implicit incompressible SPH method with a pressure poisson equation made by the combination of a symmetric SPH pressure force and a SPH discretization of the continuity equation.
    The pressure poisson equation is used to compute the pressure so that the pressure forces correct the intermediate velocities to a divergence-free state.

    A predictive corrective incompressible SPH method was proposed by Solenthaler et al.~\cite{solenthaler} based on the Lagrangian SPH model.
    Incompressibility was enforced using a prediction correction scheme to determine the particle pressures.
    To achieve this the information about the density fluctuation was actively propagated through the fluid and the pressure values where updated until the targeted density was satisfied.

    %A ghost fluid approach for free surface and solid boundary conditions for a SPH fluid simulation what proposed by Schechter et al.~\cite{ghost}.

@@

% Should be mentioned early since it is used in mostly of the algorithms
\subsection{Kernel}
    A kernel function is used to simulate how particle-particle interactions decrease with the distance between the current particle and its neighbours. In SPH simulations this is an approximation of the Gaussian kernel function. Different kernels have been tested in previous works i.e. the poly6 kernel, the spiky kernel and the cubic spline kernel. According to Bender et. al.~\cite{bender} the cubic spline kernel presented by Monaghan~\cite{monaghan} was used. The kernel is described by Equation~\ref{eq:kernel}, where $q(x)=\frac{\left \| x \right \|}{h}$, $x$ is the distance between the current particle and a neighbour particle and $h$ is the support radius for the kernel. Particles further away than the support radius will not affect the current particle.

@@

\subsection{Divergence solver}
\subsection{Density solver}

\subsection{Navier-stokes}
% Write about non pressure forces, velocity osv..?
\subsection{Adapted time step}

% CFL condition, explain why its necessary and how its assured that each rendered frame has the same time step.
In order to assure a stable simulation it is necessary to adapt the time step depending on the motion. 
The quicker the particles move the smaller time steps are taken such that the numerical 
error of the explicit time integration is negligible. For this is a condition proposed by Courant–Friedrichs–Lewy used often refered as the CFL-condition show in equation \ref{eq:cfl}. The time step $\Delta t$ is equal to a fraction of the particle's diameter $d$ divided by the maximum velocity $\vect{V_{max}}$ in the scene.

@@

\subsection{Density and alpha factors}
\subsection{Screen space fluid rendering}
